(Deep) Induction Rules for GADTs\vspace*{-0.1in}

### 2021-08-17 induction-paper/revised-induction-paper.tex
revised induction paper

It is not possible to extend structural induction to {\em truly}
nested types, i.e., to nested types whose recursive occurrences appear
below themselves. The quintessential example of such a type is that of
bushes\footnote{To define truly nested types in Agda we must use the
  $\mathsf{NO\_POSITIVITY\_CHECK}$ flag. A similar flag is required in
  Coq.}\cite{bm98}:

@@

@@

In this section we use deep induction for the $\mathsf{LTerm}$ GADT
from~\eqref{eq:eq_lterm} to extract the type from a lambda term.  The
following predicate either returns the type of its input lambda term
if that type can be inferred or indicates that the type inference
fails:

@@

\begin{align*}
  &\mathsf{GetType : \forall \, (A : Set) \to LTerm\,A \to Set} \\
  &\mathsf{GetType \,A \,t = Maybe \, (LType \, A)}
\end{align*}
The predicate $\mathsf{GetType}$ uses the standard $\mathsf{Maybe}$
data type to represent failure of type inference. It is defined by:\looseness=-1

>>>>

It is not possible to extend structural induction to {\em truly}
nested types, i.e., to nested types whose recursive occurrences appear
below themselves. The quintessential example of such a type is that of
bushes\footnote{To define truly nested types in Agda we must use the
  $\mathsf{NO\_POSITIVITY\_CHECK}$ flag, and to define functions over
  them we must use the $\mathsf{TERMINATING}$ flag.  (Similar flags
  are required in Coq.) Although as programmers we know from the
  metatheory in~\cite{jp19} that $\mathsf{Bush}$ is well-defined and
  the functions we define over them terminate, the flags are necessary
  because Agda fails to infer these facts. Analogous comments apply at
  several places below.}\cite{bm98}:

@@

Under some circumstances deep induction can be mimicked by
hand-threading applications of structural induction through the layers
of data comprising a deep data type. But this is not the case when,
e.g., one or more of the custom predicates in the data type's deep induction
rule cannot be defined by structural induction.

@@

In this section we use deep induction for the $\mathsf{LTerm}$ GADT
from~\eqref{eq:eq_lterm} to infer the type from a lambda term. The
following predicate either returns the type of its input lambda term
if that type can be inferred or indicates that the type inference
fails:

@@

\begin{align*}
  &\mathsf{GetType : \forall \, (A : Set) \to LTerm\,A \to Set} \\
  &\mathsf{GetType \,A \,t = Maybe \, (LType \, A)}
\end{align*}
Of course, since $\mathsf{GetType}$ is (trivially) defined by
structural induction, we could perform type inference using
hand-threaded applications of structural induction as observed at the
end of Section~\ref{sec:ADTs-and-nesteds}. Nevertheless, the example
as given nicely illustrates deep induction.

By construction every lambda term in $\mathsf{LTerm}$ is well-typed,
although that (necessarily unique) type cannot always be inferred.
The predicate $\mathsf{GetType}$ uses the standard $\mathsf{Maybe}$
data type to represent failure of type inference. It is defined
by:\looseness=-1
